---
title: "Plots - use-wear archaeology"
author: "Lisa Schunk"
date: "`r Sys.time()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cerulean
    highlight: pygments
    number_sections: true
  github_document: 
    toc: true
    toc_depth: 3
    html_preview: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all", 
  knit_root_dir = rprojroot::find_rstudio_root_file()) })
---

```{r Knitr Options, include=FALSE}
knitr::opts_chunk$set(comment=NA, message=FALSE, indent="", error=TRUE)
````


# Goal of the script
This script plots all variables to see which ones should be used for further analysis.  
Scatterplot of each variable will be plotted.   



```{r}
dir_in <- "analysis/derived_data/"
dir_out <- "analysis/plots"
```

Raw data must be located in `r paste0("~/", dir_in)`.  
Formatted data will be saved in `r paste0("~/", dir_out)`.
The knit directory for this script is the project directory.

---


# Load packages
```{r}
library(R.utils)
library(ggplot2)
library(tools)
library(tidyverse)
library(patchwork)
library(doBy)
library(ggrepel)
library(openxlsx)
library(wesanderson)
```


---

# Get name, path and information of the file 
```{r}
data_file <- list.files(dir_in, pattern = "\\.Rbin$", full.names = TRUE)
md5_in <- md5sum(data_file)
info_in <- data.frame(file = basename(names(md5_in)), checksum = md5_in, row.names = NULL)
```


The checksum (MD5 hashes) of the imported file is:  
```{r, echo = FALSE}
info_in
```


# Load data into R object
```{r}
imp_data <- loadObject(data_file)
str(imp_data)
```

The imported file is: "`r paste0("~/", data_file)`"  

---


# Prepare variables
## Define numeric variables
```{r}
num.var <- 24:length(imp_data)
```

The following variables will be used: 

```{r, echo=FALSE}
for (i in num.var) cat("[",i,"] ", names(imp_data)[i], "\n", sep="")
```

---


# Plot each of the selected numeric variable 
## Plot of all samples wiht the raw material, the varaibles and the use-wear type as information
```{r}
# splits the data in the individual 35 samples
imp_data[["Sample_material"]] <- paste(imp_data$Raw.material, imp_data$Sample, sep = " ")

sp <- split(imp_data, imp_data[["Site"]])

usewear <- levels(imp_data$Usewear.type)


for (i in num.var){
  
  # gets the min/max range of the data set 
  range_var <- range(imp_data[[i]]) 
 
   # plot
  p <- vector(mode = "list", length = length(sp))
  names(p) <- names(sp)
  
  for (j in seq_along(sp)) {
	p[[j]] <- ggplot(data = sp[[j]], aes_string(x = "Location", y = names(imp_data)[i], colour = "Usewear.type", shape = "Sublocation")) + 
	     # avoids overplotting
	     geom_jitter(size = 3,  position = position_jitter(width = 0.4, seed = 1)) + 
	     coord_cartesian(ylim = range_var) + 
	     theme_classic() +
	     labs(colour = "Use-wear type") +
	     facet_wrap(~ Sample_material, nrow = 3) +
	     labs(y = gsub("\\.", " ", names(imp_data)[i])) +
	     scale_colour_hue(h = c(25,225),limits = levels(sp[[j]][["Usewear.type"]])) +
	     if(j != 1) ylab(NULL) 
  }
  p_all <- wrap_plots(p) + plot_layout(width = c(8/13, 2/13, 3/13), guides = "collect")  
  
  print(p_all)
  
  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_plot_", 
	                   names(imp_data)[i], ".pdf")
	ggsave(filename = file_out, plot = p_all, path = dir_out, device = "pdf", width = 28, height = 20, units = "cm" )
}

```

## Boxplot of all the varaibles combined witht the use-wear type 
```{r}
# plot
for (i in num.var){
  
  p2 <- ggplot(data = imp_data, aes_string(x = "Usewear.type", y = names(imp_data)[i],  fill = "Tool.type")) +
        geom_boxplot() +
        theme_classic() +
        labs( x = "Use-wear type", title = " ") +
        labs(y = gsub("\\.", " ", names(imp_data)[i])) +
        labs(fill = "Tool type") +
        scale_fill_manual(values=wes_palette(n = 4,name = "Royal1"))
        
  print(p2)

  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_boxplot_", 
	                   names(imp_data)[i], ".pdf")
	ggsave(filename = file_out, plot = p2, path = dir_out, device = "pdf",width = 20, height = 25, units = "cm")
      
}  

```

## Scatterplot of selected varaibles combined witht the use-wear type 
```{r}
# plot
  p3 <- ggplot(data = imp_data, mapping = aes(x = Sa, y = Sq, colour = Usewear.type)) +  
        #geom_point(mapping = aes(x = Sa, y = Sq, colour = Usewear.type)) +
        #geom_jitter(position=position_jitter(2)) + 
        theme_classic() +
        labs(fill = "Use-wear type") +
        scale_colour_hue(h = c(25,230)) +
        geom_jitter(width = 0.1)
        
  print(p3)

  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_scatterplot_", 
	                   names(imp_data)[i], ".pdf")
	ggsave(filename = file_out, plot = p3, path = dir_out, device = "pdf")

	      
```


The files will be saved as "`r paste0("~/", dir_out, ".[ext]")`".

---


# Show plot files information
```{r}
info_out <- list.files(path = dir_out, pattern = "\\.pdf$", 
                       full.names = TRUE) %>% 
            md5sum()
```


The checksum (MD5 hashes) of the exported files are:  
```{r, echo = FALSE}
info_out
```



---

# sessionInfo() and RStudio version
```{r}
sessionInfo()
```

RStudio version `r readLines("analysis/scripts/RStudioVersion.txt", n = 1)`.


---

END OF SCRIPT

