---
title: "Plots - use-wear archaeology"
author: "Lisa Schunk"
date: "`r Sys.time()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cerulean
    highlight: pygments
    number_sections: true
  github_document: 
    toc: true
    toc_depth: 3
    html_preview: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all", 
  knit_root_dir = rprojroot::find_rstudio_root_file()) })
---

```{r Knitr Options, include=FALSE}
knitr::opts_chunk$set(comment=NA, message=FALSE, indent="", error=TRUE)
````


# Goal of the script
This script plots all variables to see which ones should be used for further analysis.  
Scatterplot of each variable will be plotted.   



```{r}
dir_in <- "analysis/derived_data/"
dir_out <- "analysis/plots"
```

Raw data must be located in `r paste0("~/", dir_in)`.  
Formatted data will be saved in `r paste0("~/", dir_out)`.
The knit directory for this script is the project directory.

---


# Load packages
```{r}
library(R.utils)
library(ggplot2)
library(tools)
library(tidyverse)
library(patchwork)
library(doBy)
library(ggrepel)
library(openxlsx)
library(wesanderson)
library(ggfortify)

```


---

# Get name, path and information of the file 
```{r}
data_file <- list.files(dir_in, pattern = "\\.Rbin$", full.names = TRUE)
md5_in <- md5sum(data_file)
info_in <- data.frame(file = basename(names(md5_in)), checksum = md5_in, row.names = NULL)
```


The checksum (MD5 hashes) of the imported file is:  
```{r, echo = FALSE}
info_in
```


# Load data into R object
```{r}
imp_data <- loadObject(data_file)
str(imp_data)
```

The imported file is: "`r paste0("~/", data_file)`"  

---


# Prepare variables
## Define numeric variables
```{r}
num.var <- 24:length(imp_data)
```

The following variables will be used: 

```{r, echo=FALSE}
for (i in num.var) cat("[",i,"] ", names(imp_data)[i], "\n", sep="")
```

---


# Plot each of the selected numeric variables 
## Plot of all samples with raw material,  variables and use-wear type as information
```{r}
# splits the data in the individual 35 samples
imp_data[["Sample_material"]] <- paste(imp_data$Raw.material, imp_data$Sample, sep = " ")

sp <- split(imp_data, imp_data[["Site"]])

usewear <- levels(imp_data$Usewear.type)


for (i in num.var){
  
  # gets the min/max range of the data set 
  range_var <- range(imp_data[[i]]) 
 
   # plot
  p <- vector(mode = "list", length = length(sp))
  names(p) <- names(sp)
  
  for (j in seq_along(sp)) {
	p[[j]] <- ggplot(data = sp[[j]], aes_string(x = "Location", y = names(imp_data)[i],
	                                            colour = "Usewear.type", shape =
	                                           "Sublocation")) + 
	     # avoids overplotting
	     geom_jitter(size = 3,  position = position_jitter(width = 0.4, seed = 1)) + 
	     coord_cartesian(ylim = range_var) + 
	     theme_classic() +
	     labs(colour = "Use-wear type") +
	     facet_wrap(~ Sample_material, nrow = 3) +
	     labs(y = gsub("\\.", " ", names(imp_data)[i])) +
	     scale_colour_hue(h = c(25,225),limits = levels(sp[[j]][["Usewear.type"]])) +
	     if(j != 1) ylab(NULL) 
  }
  p_all <- wrap_plots(p) + plot_layout(width = c(8/13, 2/13, 3/13), guides = "collect")  
  
  print(p_all)
  
  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_plot_", 
	                   names(imp_data)[i], ".pdf")
	ggsave(filename = file_out, plot = p_all, path = dir_out, device = "pdf", width = 38,
	       height = 26, units = "cm" )
}

```


## Boxplot of all the variables combined with the use-wear type 
```{r}
# plot
for (i in num.var){
  
  p2 <- ggplot(data = imp_data, aes_string(x = "Usewear.type", y = names(imp_data)[i],
                                           fill = "Tool.type")) +
        geom_boxplot() +
        theme_classic() +
        labs( x = "Use-wear type", title = " ") +
        labs(y = gsub("\\.", " ", names(imp_data)[i])) +
        labs(fill = "Tool type") +
        scale_fill_manual(values=wes_palette(n = 4, name = "Royal1"))
        
  print(p2)

  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_boxplot_", 
	                   names(imp_data)[i], ".pdf")
	ggsave(filename = file_out, plot = p2, path = dir_out, device = "pdf", width = 17,
	       height = 25, units = "cm")
      
}  

```


## Boxplot of all the variables combined with the use-wear type - tool types separated
```{r}
# Keilmesser
# sorts the data according to the technological class 
sort_data <- imp_data[ , ] %>% arrange(Tool.type)
# adds indices as row names 
row.names(sort_data) <- 1:nrow(sort_data)
# excludes all other tool types  
KM_data <- sort_data [1:111, ]

for (i in num.var){
  
  KM <- ggplot(data = KM_data, aes_string(x = "Usewear.type", y = names(KM_data)[i],
                                           fill = "Tool.type")) +
        geom_boxplot() +
        theme_classic() +
        labs( x = "Use-wear type", title = " ") +
        labs(y = gsub("\\.", " ", names(KM_data)[i])) +
        labs(fill = "Tool type") +
        scale_fill_manual(values = wes_palette(n = 4, name = "Royal1"))
        
  print(KM)

  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_KM_boxplot", 
	                   names(KM_data)[i], ".pdf")
	ggsave(filename = file_out, plot = KM, path = dir_out, device = "pdf", width = 17,
	       height = 25, units = "cm")
      
}  


# Pradnick scraper 
# excludes all other tool types  
PS_data <- sort_data [112:120, ]

for (i in num.var){
  
  PS <- ggplot(data = PS_data, aes_string(x = "Usewear.type", y = names(PS_data)[i],
                                           fill = "Tool.type")) +
        geom_boxplot() +
        theme_classic() +
        labs( x = "Use-wear type", title = " ") +
        labs(y = gsub("\\.", " ", names(PS_data)[i])) +
        labs(fill = "Tool type") +
        scale_fill_manual(values = "#C93312")
         
  print(PS)

  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_PS_boxplot", 
	                   names(PS_data)[i], ".pdf")
	ggsave(filename = file_out, plot = PS, path = dir_out, device = "pdf", width = 17,
	       height = 25, units = "cm")
      
}  


# Scraper 
# excludes all other tool types  
S_data <- sort_data [133:150, ]

for (i in num.var){
  
  S <- ggplot(data = S_data, aes_string(x = "Usewear.type", y = names(S_data)[i],
                                           fill = "Tool.type")) +
        geom_boxplot() +
        theme_classic() +
        labs( x = "Use-wear type", title = " ") +
        labs(y = gsub("\\.", " ", names(S_data)[i])) +
        labs(fill = "Tool type") +
        scale_fill_manual(values = "#DC863B")
         
  print(S)

  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_S_boxplot", 
	                   names(S_data)[i], ".pdf")
	ggsave(filename = file_out, plot = S, path = dir_out, device = "pdf", width = 17,
	       height = 25, units = "cm")
      
}  


# Pradnick spall 
# excludes all other tool types  
LSS_data <- sort_data [121:132, ]

for (i in num.var){
  
  LSS <- ggplot(data = LSS_data, aes_string(x = "Usewear.type", y = names(LSS_data)[i],
                                           fill = "Tool.type")) +
        geom_boxplot() +
        theme_classic() +
        labs( x = "Use-wear type", title = " ") +
        labs(y = gsub("\\.", " ", names(LSS_data)[i])) +
        labs(fill = "Tool type") +
        scale_fill_manual(values =  "#FAEFD1")
         
  print(LSS)

  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_LSS_boxplot", 
	                   names(LSS_data)[i], ".pdf")
	ggsave(filename = file_out, plot = LSS, path = dir_out, device = "pdf", width = 17,
	       height = 25, units = "cm")
      
}  

```


## Scatterplots of selected variables combined with the use-wear type 
```{r}
# plot 
# plots Sa against Sq
p3 <- ggplot(data = imp_data) +  
      geom_point(mapping = aes(x = Sa, y = Sq, colour = Usewear.type)) +
      theme_classic() +
      labs(colour = "Use-wear type") +
      scale_colour_hue(h = c(25, 230)) 
        
print(p3)

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_SA_scatterplot_", ".pdf")
ggsave(filename = file_out, plot = p3, path = dir_out, device = "pdf")

 
# plots epLsar against Asfc 
p4 <- ggplot(data = imp_data) +  
      geom_point(mapping = aes(x = Asfc, y = epLsar, colour = Usewear.type)) +
      theme_classic() +
      labs(colour = "Use-wear type") +
      scale_colour_hue(h = c(25, 230)) 
        
print(p4) 

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_Asfc_scatterplot_", ".pdf")
ggsave(filename = file_out, plot = p4, path = dir_out, device = "pdf")

  
# plots Sq against Vmc 
p5 <- ggplot(data = imp_data) +  
      geom_point(mapping = aes(x = Sq, y = Vmc, colour = Usewear.type)) +
      theme_classic() +
      labs(colour = "Use-wear type") +
      scale_colour_hue(h = c(25, 230)) 
       
print(p5)

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_Sq_scatterplot_", ".pdf")
ggsave(filename = file_out, plot = p5, path = dir_out, device = "pdf")


# plots Mean depth of furrows against mean density of furrows  
p6 <- ggplot(data = imp_data) +  
      geom_point(mapping = aes(x = Mean.depth.of.furrows, y = Mean.density.of.furrows,
                               colour = Usewear.type)) +
      theme_classic() +
      labs(colour = "Use-wear type") +
      scale_colour_hue(h = c(25, 230)) +
      labs(x = "Mean depth of furrows", y = "Mean density of furrows")
  
print(p6)

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_furrows_scatterplot_", ".pdf")
ggsave(filename = file_out, plot = p6, path = dir_out, device = "pdf")
  


```


## Principal component analysis 
```{r}
imp_data.pca <- prcomp(imp_data[24:30], scale. = TRUE) 

imp_data.pca$x
pca <- cbind(imp_data[, 1:9], imp_data.pca$x)

# Using ggfortigy
a<- autoplot(imp_data.pca, data = pca, colour = "Usewear.type",
         label = TRUE, label.size = 2,
         loadings = TRUE, loadings.label = TRUE, loadings.label.size  = 3) # colour should be categorical

print(a)

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_factor_analysis", ".pdf")
ggsave(filename = file_out, plot = a, path = dir_out, device = "pdf")
  

# using ggplot
# PCA use-wear type
b <- ggplot(pca, aes(PC1, PC2, col = Tool.type, fill = Usewear.type)) +
     stat_ellipse(geom = "polygon", col = "black", alpha = 0.5) +
     scale_fill_manual(values = wes_palette(n = 11, name = "Royal1", type = "continuous")) + 
     geom_point(shape = 21, col = "black") +
     labs(fill = "Use-wear type") +
     theme_classic()

print(b)

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_PCA_Usewear", ".pdf")
ggsave(filename = file_out, plot = b, path = dir_out, device = "pdf")

# PCA tool type 
c <- ggplot(pca, aes(PC1, PC2, col = Tool.type, fill = Tool.type)) +
          stat_ellipse(geom = "polygon", col = "black", alpha = 0.5) +
          scale_fill_manual(values = wes_palette(n = 4, name = "FantasticFox1")) +
          geom_point(shape = 21, col = "black") +
          labs(fill = "Tool type") +
          theme_classic()

print(c)

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_PCA_Tool_type", ".pdf")
ggsave(filename = file_out, plot = c, path = dir_out, device = "pdf")


```


The files will be saved as "`r paste0("~/", dir_out, ".[ext]")`".

---


# Show plot files information
```{r}
info_out <- list.files(path = dir_out, pattern = "\\.pdf$", 
                       full.names = TRUE) %>% 
            md5sum()
```


The checksum (MD5 hashes) of the exported files are:  
```{r, echo = FALSE}
info_out
```



---

# sessionInfo() and RStudio version
```{r}
sessionInfo()
```

RStudio version `r readLines("analysis/scripts/RStudioVersion.txt", n = 1)`.


---

END OF SCRIPT

