---
title: "Plots - use-wear archaeology"
author: "Lisa Schunk"
date: "`r Sys.time()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cerulean
    highlight: pygments
    number_sections: true
  github_document: 
    toc: true
    toc_depth: 3
    html_preview: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all", 
  knit_root_dir = rprojroot::find_rstudio_root_file()) })
---

```{r Knitr Options, include=FALSE}
knitr::opts_chunk$set(comment=NA, message=FALSE, indent="", error=TRUE)
````


# Goal of the script
This script plots all variables to see which ones should be used for further analysis.  
Scatterplot of each variable will be plotted.   



```{r}
dir_in <- "analysis/derived_data/"
dir_out <- "analysis/plots"
```

Raw data must be located in `r paste0("~/", dir_in)`.  
Formatted data will be saved in `r paste0("~/", dir_out)`.
The knit directory for this script is the project directory.

---


# Load packages
```{r}
library(R.utils)
library(ggplot2)
library(tools)
library(tidyverse)
library(patchwork)
library(doBy)
library(ggrepel)
library(openxlsx)
library(wesanderson)
library(ggfortify)

```


---

# Get name, path and information of the file 
```{r}
data_file <- list.files(dir_in, pattern = "\\.Rbin$", full.names = TRUE)
md5_in <- md5sum(data_file)
info_in <- data.frame(file = basename(names(md5_in)), checksum = md5_in, row.names = NULL)
```


The checksum (MD5 hashes) of the imported file is:  
```{r, echo = FALSE}
info_in
```


# Load data into R object
```{r}
imp_data <- loadObject(data_file)
str(imp_data)
```

The imported file is: "`r paste0("~/", data_file)`"  

---


# Prepare variables
## Define numeric variables
```{r}
num.var <- 24:length(imp_data)
```

The following variables will be used: 

```{r, echo=FALSE}
for (i in num.var) cat("[",i,"] ", names(imp_data)[i], "\n", sep="")
```

---




# Plot each of the selected numeric variables 
## Colour definitions for use-wear types
```{r}
#05100c black 
#999999 grey 
#52854c green 
#c3d7a4 light green 
#487bb6 blue 
#a6cee3 light blue 
#9a0f0f red
#d16103 orange
#fdbf6f apricot
#ffdb6d yellow
#985633 brown 
#134680 dark blue


custom.col <- data.frame(type = levels(imp_data$Usewear.typ), 
                         col = c("#999999", "#52854c", "#c3d7a4", "#487bb6", "#9a0f0f", "#fdbf6f",
                                 "#d16103", "#ffdb6d", "#985633", "#134680", "#05100c")) 


```



## Plot of all samples with raw material,  variables and use-wear type as information
```{r}
# splits the data in the individual 35 samples
imp_data[["Sample_material"]] <- paste(imp_data$Raw.material, imp_data$Sample, sep = " ")

sp <- split(imp_data, imp_data[["Site"]])

usewear <- levels(imp_data$Usewear.type)


for (i in num.var){
  
  # gets the min/max range of the data set 
  range_var <- range(imp_data[[i]]) 
 
   # plot
  p <- vector(mode = "list", length = length(sp))
  names(p) <- names(sp)
  
  for (j in seq_along(sp)) {
    col_j <- custom.col[custom.col$type %in% levels(factor(sp[[j]][["Usewear.type"]])), "col"]

	p[[j]] <- ggplot(data = sp[[j]], aes_string(x = "Location", y = names(imp_data)[i],
	                                            colour = "Usewear.type", shape =
	                                           "Sublocation")) + 
	     # avoids overplotting
	     geom_jitter(size = 3,  position = position_jitter(width = 0.4, seed = 1)) + 
	     coord_cartesian(ylim = range_var) + 
	     theme_classic() +
	     labs(colour = "Use-wear type") +
	     facet_wrap(~ Sample_material, nrow = 3) +
	     labs(y = gsub("\\.", " ", names(imp_data)[i])) +
	     scale_colour_manual(values = col_j) +
	     theme(text = element_text(size = 23)) + 
	     if(j != 1) ylab(NULL) 

  }
  p_all <- wrap_plots(p) + plot_layout(width = c(8/13, 2/13, 3/13), guides = "collect")  
  
  print(p_all)
  
  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_plot_", 
	                   names(imp_data)[i], ".pdf")
	ggsave(filename = file_out, plot = p_all, path = dir_out, device = "pdf", width = 65,
	       height = 53, units = "cm" )
}

```


## Boxplot of all the variables combined with the use-wear type (without outliers)
```{r}

# Wes Anderson color palettes Rushmore = c("#E1BD6D", "#EABE94", "#0B775E", "#35274A" ,"#F2300F")
custom.col2 <- data.frame(type = levels(imp_data$Tool.typ), 
                         col = c( "#0B775E", "#E1BD6D", "#F2300F", "#35274A")) 
imp_data$col <- custom.col2[imp_data$Tool.typ, "col"]


# excludes the outliers 
# adds the indices as row numbers 
imp_data <- imp_data %>% mutate(id = row_number())
imp_data2 <- imp_data[-c(55, 63, 115, 116), ]


# plot
for (i in num.var){
  
  p2 <- ggplot(data = imp_data2, aes_string(x = "Usewear.type", y = names(imp_data)[i],
                                           fill = "Tool.type")) +
        geom_boxplot() +
        scale_fill_manual(values = custom.col2$col)+
        theme_classic() +
        labs( x = "Use-wear type", title = " ") +
        labs(y = gsub("\\.", " ", names(imp_data)[i])) +
        labs(fill = "Tool type") 
       
  print(p2)

  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_boxplot_", 
	                   names(imp_data)[i], ".pdf")
	ggsave(filename = file_out, plot = p2, path = dir_out, device = "pdf", width = 17,
	       height = 25, units = "cm")
      
}  

```


## Boxplot of all the variables combined with the use-wear type - tool types separated (without outliers)
```{r}
# Keilmesser
# sorts the data according to the technological class 
sort_data <- imp_data2[ , ] %>% arrange(Tool.type)
# adds indices as row names 
row.names(sort_data) <- 1:nrow(sort_data)
# excludes all other tool types  
KM_data <- sort_data [1:107, ]

for (i in num.var){
  
  KM <- ggplot(data = KM_data, aes_string(x = "Usewear.type", y = names(KM_data)[i],
                                           fill = "Tool.type")) +
        geom_boxplot() +
        theme_classic() +
        labs( x = "Use-wear type", title = " ") +
        labs(y = gsub("\\.", " ", names(KM_data)[i])) +
        labs(fill = "Tool type") +
        scale_fill_manual(values = "#0B775E")
        
  print(KM)

  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_KM_boxplot", 
	                   names(KM_data)[i], ".pdf")
	ggsave(filename = file_out, plot = KM, path = dir_out, device = "pdf", width = 17,
	       height = 25, units = "cm")
      
}  


# Pradnick scraper 
# excludes all other tool types  
PS_data <- sort_data [108:116, ]

for (i in num.var){
  
  PS <- ggplot(data = PS_data, aes_string(x = "Usewear.type", y = names(PS_data)[i],
                                           fill = "Tool.type")) +
        geom_boxplot() +
        theme_classic() +
        labs( x = "Use-wear type", title = " ") +
        labs(y = gsub("\\.", " ", names(PS_data)[i])) +
        labs(fill = "Tool type") +
        scale_fill_manual(values = "#E1BD6D")
         
  print(PS)

  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_PS_boxplot", 
	                   names(PS_data)[i], ".pdf")
	ggsave(filename = file_out, plot = PS, path = dir_out, device = "pdf", width = 17,
	       height = 25, units = "cm")
      
}  


# Scraper 
# excludes all other tool types  
S_data <- sort_data [129:146, ]

for (i in num.var){
  
  S <- ggplot(data = S_data, aes_string(x = "Usewear.type", y = names(S_data)[i],
                                           fill = "Tool.type")) +
        geom_boxplot() +
        theme_classic() +
        labs( x = "Use-wear type", title = " ") +
        labs(y = gsub("\\.", " ", names(S_data)[i])) +
        labs(fill = "Tool type") +
        scale_fill_manual(values = "#35274A")
         
  print(S)

  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_S_boxplot", 
	                   names(S_data)[i], ".pdf")
	ggsave(filename = file_out, plot = S, path = dir_out, device = "pdf", width = 17,
	       height = 25, units = "cm")
      
}  


# Pradnick spall 
# excludes all other tool types  
LSS_data <- sort_data [117:128, ]

for (i in num.var){
  
  LSS <- ggplot(data = LSS_data, aes_string(x = "Usewear.type", y = names(LSS_data)[i],
                                           fill = "Tool.type")) +
        geom_boxplot() +
        theme_classic() +
        labs( x = "Use-wear type", title = " ") +
        labs(y = gsub("\\.", " ", names(LSS_data)[i])) +
        labs(fill = "Tool type") +
        scale_fill_manual(values =  "#F2300F")
         
  print(LSS)

  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_LSS_boxplot", 
	                   names(LSS_data)[i], ".pdf")
	ggsave(filename = file_out, plot = LSS, path = dir_out, device = "pdf", width = 17,
	       height = 25, units = "cm")
      
}  

```


## Histogram of the use-wear types (without outliers)
```{r}
custom.col <- data.frame(type = levels(imp_data$Usewear.typ), 
                         col = c("#999999", "#52854c", "#c3d7a4", "#487bb6", "#9a0f0f", "#fdbf6f",
                                 "#d16103", "#ffdb6d", "#985633", "#134680", "#05100c")) 

col <- custom.col[custom.col$type %in% levels(imp_data[["Usewear.type"]]), "col"]



# plot
for (i in num.var){
  
  
  p_use <- ggplot(data = imp_data2, aes_string(x = names(imp_data)[i])) +
           geom_histogram(bins = 15, aes(fill = Usewear.type)) +
           theme_classic() +
           labs(x = gsub("\\.", " ", names(imp_data)[i])) +
           labs(fill = "Usewear type", y = NULL) +
           facet_wrap(~Usewear.type)+
           scale_fill_manual(values =  col)
  
  print(p_use)

  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_histogram_UW_Type_", 
	                   names(imp_data)[i], ".pdf")
	ggsave(filename = file_out, plot = p_use, path = dir_out, device = "pdf")
      
}  


```


## Scatterplots of selected variables combined with the use-wear type (without outliers)
```{r}

custom.col <- data.frame(type = levels(imp_data$Usewear.typ), 
                         col = c("#999999", "#52854c", "#c3d7a4", "#487bb6", "#9a0f0f", "#fdbf6f",
                                 "#d16103", "#ffdb6d", "#985633", "#134680", "#05100c")) 

col <- custom.col[custom.col$type %in% levels(imp_data[["Usewear.type"]]), "col"]

# plot 
# plots Sa against Sq
p3 <- ggplot(data = imp_data2) +  
      geom_point(mapping = aes(x = Sa, y = Sq, colour = Usewear.type)) +
      theme_classic() +
      labs(colour = "Use-wear type") +
      scale_colour_manual(values =  col)
        
print(p3)

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_SA_scatterplot_", ".pdf")
ggsave(filename = file_out, plot = p3, path = dir_out, device = "pdf")

 
# plots epLsar against Asfc 
p4 <- ggplot(data = imp_data2) +  
      geom_point(mapping = aes(x = Asfc, y = epLsar, colour = Usewear.type)) +
      theme_classic() +
      labs(colour = "Use-wear type") +
       scale_colour_manual(values =  col)
        
print(p4) 

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_Asfc_scatterplot_", ".pdf")
ggsave(filename = file_out, plot = p4, path = dir_out, device = "pdf")

  
# plots Sq against Vmc 
p5 <- ggplot(data = imp_data2) +  
      geom_point(mapping = aes(x = Sq, y = Vmc, colour = Usewear.type)) +
      theme_classic() +
      labs(colour = "Use-wear type") +
       scale_colour_manual(values =  col)
       
print(p5)

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_Sq_scatterplot_", ".pdf")
ggsave(filename = file_out, plot = p5, path = dir_out, device = "pdf")


# plots Mean depth of furrows against mean density of furrows  
p6 <- ggplot(data = imp_data2) +  
      geom_point(mapping = aes(x = Mean.depth.of.furrows, y = Mean.density.of.furrows,
                               colour = Usewear.type)) +
      theme_classic() +
      labs(colour = "Use-wear type") +
       scale_colour_manual(values =  col) +
      labs(x = "Mean depth of furrows", y = "Mean density of furrows")
  
print(p6)

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_furrows_scatterplot_", ".pdf")
ggsave(filename = file_out, plot = p6, path = dir_out, device = "pdf")
  

```


## Scatterplots of selected variables combined with the two tool types: Keilmesser and Pradnick scraper (without outliers)
```{r}
# selects only Keilmesser and Pradnick scraper
KM_PS <- filter(imp_data2, Tool.type == "Keilmesser" | Tool.type == "Pradnick scraper") 


custom.col2b <- data.frame(type = unique(KM_PS$Tool.type), 
                         col = c("#0B775E", "#E1BD6D")) 

col2b <- custom.col2b[custom.col2b$type %in% unique(KM_PS[["Tool.type"]]), "col"]

# plot 
# plots Sa against Sq
p7 <- ggplot(data = KM_PS) +  
      geom_point(mapping = aes(x = Sa, y = Sq, colour = Tool.type)) +
      theme_classic() +
      labs(colour = "Tool type") +
      scale_colour_manual(values =  col2b)
        
print(p7)

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_SA_KM.PS_scatterplot_", ".pdf")
ggsave(filename = file_out, plot = p7, path = dir_out, device = "pdf")

 
# plots epLsar against Asfc 
p8 <- ggplot(data = KM_PS) +  
      geom_point(mapping = aes(x = Asfc, y = epLsar, colour = Tool.type)) +
      theme_classic() +
      labs(colour = "Tool type") +
      scale_colour_manual(values =  col2b)
        
print(p8) 

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_Asfc_KM.PS_scatterplot_", ".pdf")
ggsave(filename = file_out, plot = p8, path = dir_out, device = "pdf")

  
# plots Sq against Vmc 
p9 <- ggplot(data = KM_PS) +  
      geom_point(mapping = aes(x = Sq, y = Vmc, colour = Tool.type)) +
      theme_classic() +
      labs(colour = "Tool type") +
      scale_colour_manual(values =  col2b)
       
print(p9)

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_Sq_KM.PS_scatterplot_", ".pdf")
ggsave(filename = file_out, plot = p9, path = dir_out, device = "pdf")


# plots Mean depth of furrows against mean density of furrows  
p10 <- ggplot(data = KM_PS) +  
       geom_point(mapping = aes(x = Mean.depth.of.furrows, y = Mean.density.of.furrows,
                               colour = Tool.type)) +
       theme_classic() +
       labs(colour = "Tool type") +
       scale_colour_manual(values =  col2b) +
       labs(x = "Mean depth of furrows", y = "Mean density of furrows")
  
print(p10)

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_furrows_KM.PS_scatterplot_", ".pdf")
ggsave(filename = file_out, plot = p10, path = dir_out, device = "pdf")
  

```


## Principal component analysis 
### PCA Use-wear type (without outliers)
```{r}
# uses for the PCA only selected variables: Sq, SSK, Vmc, Isotropy, Mean, density of furrows, Asfc, HAsfc9 
imp_data.pca <- prcomp(imp_data2[, c(24:25, 42, 47, 51, 54,56)], scale. = TRUE) 


custom.col1 <- data.frame(type = levels(imp_data$Usewear.typ), 
                         col = c("#999999", "#52854c", "#c3d7a4", "#487bb6", "#9a0f0f", "#fdbf6f",
                                 "#d16103", "#ffdb6d", "#985633", "#134680", "#05100c")) 
imp_data$col <- custom.col1[imp_data$Usewear.typ, "col"]

# Using ggfortify
a<- autoplot(imp_data.pca, data = imp_data2, colour = "Usewear.type", size = 2,
             loadings = TRUE, loadings.colour = "black", loadings.label = TRUE, loadings.label.colour = "black", 
             loadings.label.size  = 4, loadings.label.hjust = 1, loadings.label.vjust = 1,  
             frame = TRUE, frame.type = "convex", frame.colour = "Usewear.type", frame.alpha = 0) + 
             theme_classic() +
             scale_colour_manual(values = custom.col1$col)
         
print(a)


# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_PCA_usewear_type", ".pdf")
ggsave(filename = file_out, plot = a, path = dir_out, device = "pdf")
  

```


### PCA Tool type (without outliers)
```{r}
# uses for the PCA only selected variables: Sq, SSK, Vmc, Isotropy, Mean density of furrows, Asfc, HAsfc9 
imp_data.pca <- prcomp(imp_data2[, c(24:25, 42, 47, 51, 54,56)], scale. = TRUE) 

# Wes Anderson color palettes Rushmore = c("#E1BD6D", "#EABE94", "#0B775E", "#35274A" ,"#F2300F")
custom.col2 <- data.frame(type = levels(imp_data$Tool.typ), 
                         col = c( "#0B775E", "#E1BD6D", "#F2300F", "#35274A")) 
imp_data$col <- custom.col2[imp_data$Tool.typ, "col"]

b <- autoplot(imp_data.pca, data = imp_data2, colour = "Tool.type", size = 2,
             loadings = TRUE, loadings.colour = "black", loadings.label = TRUE, loadings.label.colour = "black",                
             loadings.label.size  = 4, loadings.label.hjust = 1, loadings.label.vjust = 1,  
             frame = TRUE, frame.type = "convex", frame.colour = "Tool.type", frame.alpha = 0) + 
             theme_classic() +
             scale_colour_manual(values = custom.col2$col)
print(b)

# saves the plot
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_PCA_tool_tpye", ".pdf")
ggsave(filename = file_out, plot = b, path = dir_out, device = "pdf")


```




The files will be saved as "`r paste0("~/", dir_out, ".[ext]")`".

---


# Show plot files information
```{r}
info_out <- list.files(path = dir_out, pattern = "\\.pdf$", 
                       full.names = TRUE) %>% 
            md5sum()
```


The checksum (MD5 hashes) of the exported files are:  
```{r, echo = FALSE}
info_out
```



---

# sessionInfo() and RStudio version
```{r}
sessionInfo()
```

RStudio version `r readLines("analysis/scripts/RStudioVersion.txt", n = 1)`.


---

END OF SCRIPT

