---
title: "Import CSV from ConfoMap ISO25178 - use-wear archaeology"
author: "Ivan Calandra & Lisa Schunk"
date: "`r Sys.time()`"
output: word_document
---


```{r Knitr Options, include=FALSE}
knitr::opts_chunk$set(comment = NA, message = FALSE, indent = "", error = TRUE)

```

---


# Goal of the script
This script formats the output of the resulting CSV-files from applying a template
computing ISO25178 parameters in ConfoMap.
The script will:

1. Read in the original CSV-files
2. Format the data   
3. Write an XLSX-file and save an R object ready for further analysis in R 

```{r}
dir_in <- "analysis/raw_data"
dir_out <- "analysis/derived_data/"

```

Raw data must be located in `r paste0("~/", dir_in)`.  
Formatted data will be saved in `r paste0("~/", dir_out)`.

The knit directory for this script is the project directory.

---


# Load packages
```{r}
library(openxlsx)
library(tools)
library(R.utils)
library(chron)
library(tidyverse)

```


---


# Get names, path and information of all files 
```{r}
data_files <- list.files(dir_in, pattern = "\\.csv$", full.names = TRUE)
md5_in <- md5sum(data_files)
info_in <- data.frame(files = basename(names(md5_in)), checksum = md5_in, 
                      row.names = NULL)

```

The checksum (MD5 hashes) of the imported files are:  
```{r, echo = FALSE}
info_in

```


---


# Read in original CSV-files
```{r}
imp_data <- vector(mode = "list", length = length(data_files))
names(imp_data) <- basename(data_files)
# loop for import data due to the three different CSV files 
for (i in seq_along(data_files)) {
  imp_data[[i]] <- read.csv(data_files[i], header = FALSE, 
                            stringsAsFactors = FALSE, na.strings = "*****")
}
str(imp_data)

```


---


# Format data
## Merge three datasets
```{r}
# check pairwise if the three lines of headers are identical among the datasets
# merges the data based on the three lines of headers while they get only 
# used in the first CSV file 
comp <- all(sapply(list(imp_data[[1]][1:3, ], imp_data[[2]][1:3, ]), 
                   FUN = identical, imp_data[[3]][1:3, ]))
if (comp == TRUE) {
  merged_data <- rbind(imp_data[[1]], imp_data[[2]][-(1:3), ], 
                       imp_data[[3]][-(1:3), ])
} else {
  stop("The headers are not identical among the datasets")
}

str(merged_data)

```


## Keep only interesting columns and rows
```{r}
# keeps only the columns and rows of interest for the analysis 
data_keep_col <- c(1:2, 16:17, 20:22, 24:26, 29:32, 67:100)
data_keep_rows <- which(merged_data[[1]] != "#")  
data_keep <- merged_data[data_keep_rows, data_keep_col]

```


## Add headers
```{r}
head_data_keep <- unlist(merged_data[2, data_keep_col]) 
colnames(data_keep) <- gsub("\\.+", "\\.", make.names(head_data_keep))
colnames(data_keep) <- gsub("\\.$", "", colnames(data_keep)) 

```


## Identify results using frame numbers
```{r}
# combines the results from the different analysis based on the column 
# numbers (ID from MountainsMAp)
frames <- as.numeric(unlist(merged_data[1, data_keep_col]))
ID <- which(frames == 6)[-(1:2)]
ISO <- which(frames == 17)
furrow <- which(frames == 18)
diriso <- which(frames %in% 19:20)
SSFA <- which(frames %in% 21:22)

```


## Shorten the names for parameters
```{r}
# keeps only the important information of the headers 
colnames(data_keep)[ISO] <- sapply(strsplit(names(data_keep)[ISO], ".", fixed = TRUE),
                                   `[[`, 1)
colnames(data_keep)[SSFA] <- gsub("^([A-Za-z0-9]+\\.)+", "", colnames(data_keep)[SSFA])

```


## Save units
```{r}
# takes the units which were part of the headers and separates them; 
# creates a data frame
var_num <- c(ID, ISO, furrow, diriso, SSFA)
#extract 'unit' line for considered columns
units_var <- unlist(merged_data[3, data_keep_col])[var_num] 
# gets names associated to the units
names(units_var) <- head_data_keep[var_num] 
# puts all of it into a data.frame
units_var_table <- data.frame(variable = names(units_var), unit = units_var) 

```


## Convert to numeric
```{r}
for (i in var_num) {
  data_keep[[i]] <- gsub(",", "\\.", data_keep[[i]])
  data_keep[[i]] <- as.numeric(data_keep[[i]])
}

```


## Split the column 'Name' into several columns
```{r}
# these lines extract the artefact ID out of the path name
stud_name <- gsub("^([A-Za-z0-9_]+( --- ))+", "", data_keep[["Name"]])
stud_name <- gsub("([A-Za-z0-9_-]*( - ))+", "", stud_name)
split_name <- do.call(rbind, strsplit(stud_name, "_"))[, 1:3]
split_loc <- do.call(rbind, strsplit(split_name[, 2], "-"))

# splits location (A1-3,B1-3,C1-3,D1-3) in location (A,B,C,D) and sublocation (1,2,3,4)
split_subloc1 <- substr(split_loc[,1], 1, 1)
split_subloc2 <- substr(split_loc[,1], 2, 2)

# splits the ID in the separat information 
data_final <- data.frame(split_name[, -2], split_subloc1, split_subloc2, 
                         split_loc[, 2-3], data_keep[-3], stringsAsFactors = FALSE)
colnames(data_final)[1:9] <- c("Sample", "Objective", "Location", "Sublocation", 
                               "Area", "Spot", "Analysis.date", "Analysis.time",
                               "Acquisition.date.time")

```


## Format date and time columns
```{r}
data_final[["Analysis.date"]] <- as.Date(data_final[["Analysis.date"]],
                                         format = "%d.%m.%Y")
data_final[["Analysis.time"]] <- times(data_final[["Analysis.time"]])

```
The column `data_final[["Acquisition.date.time"]]` includes several formats and 
is therefore left as `character` without convertion to `POSIXct`.


## Add columns with further information and corrects 50x objectives' NAs
```{r}
# extracte the site name based on the ID
data_final[grep("R-0", data_final[["Sample"]]), "Site"] <- "Ramioul"
data_final[grep("MU-", data_final[["Sample"]]), "Site"] <- "Balve"
data_final[grep("BU-", data_final[["Sample"]]), "Site"] <- "Buhlen"
data_final[["Site"]] <- factor(data_final[["Site"]])

# adds the raw material by defining the flint samples, everything else is lydite 
data_final[["Raw.material"]] <- factor(ifelse(data_final[["Sample"]] %in% 
                                                c("MU-224", "MU-197", "R-002", "R-006",
                                                  "R-007", "R-008" ,"R-010" ,"R-013"),
                                                  "flint", "lydite"))

# add the tool type - Keilmesser
data_final[grep("MU-003", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-008", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-020", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-021", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-041", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-107", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-111", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-112", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-199", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-202", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-224", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-232", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-197", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-240", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-246", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("MU-273", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"

data_final[grep("R-002", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("R-006", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("R-007", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("R-008", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("R-020", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"

data_final[grep("BU-003", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("BU-032", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("BU-077", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"
data_final[grep("BU-173", data_final[["Sample"]]), "Tool.type"] <- "Keilmesser"

# add the tool type - Prądnik spall
data_final[grep("MU-104", data_final[["Sample"]]), "Tool.type"] <- "Prądnik spall"
data_final[grep("MU-119", data_final[["Sample"]]), "Tool.type"] <- "Prądnik spall"
data_final[grep("MU-217", data_final[["Sample"]]), "Tool.type"] <- "Prądnik spall"

data_final[grep("BU-128", data_final[["Sample"]]), "Tool.type"] <- "Prądnik spall"

# add the tool type - Scraper 
data_final[grep("MU-019", data_final[["Sample"]]), "Tool.type"] <- "Scraper"
data_final[grep("MU-025", data_final[["Sample"]]), "Tool.type"] <- "Scraper"
data_final[grep("MU-279", data_final[["Sample"]]), "Tool.type"] <- "Scraper"

data_final[grep("R-013", data_final[["Sample"]]), "Tool.type"] <- "Scraper"


# add the tool type - Prądnik scraper 
data_final[grep("R-010", data_final[["Sample"]]), "Tool.type"] <- "Prądnik scraper"
data_final[grep("BU-115", data_final[["Sample"]]), "Tool.type"] <- "Prądnik scraper"
data_final[["Tool.type"]] <- factor(data_final[["Tool.type"]])

# add the use-wear type - Balve 
data_final[data_final[["Sample"]] == "MU-003" & data_final[["Location"]] == "D" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "B"
data_final[data_final[["Sample"]] == "MU-008" & data_final[["Location"]] == "B" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "C"
data_final[data_final[["Sample"]] == "MU-020" & data_final[["Location"]] == "A" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "D"
data_final[data_final[["Sample"]] == "MU-020" & data_final[["Location"]] == "A" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "D2"
data_final[data_final[["Sample"]] == "MU-021" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "3", "Usewear.type"] <- "B"
data_final[data_final[["Sample"]] == "MU-021" & data_final[["Location"]] == "D" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "D2"
data_final[data_final[["Sample"]] == "MU-041" & data_final[["Location"]] == "D" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "E"
data_final[data_final[["Sample"]] == "MU-041" & data_final[["Location"]] == "D" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "E"
data_final[data_final[["Sample"]] == "MU-107" & data_final[["Location"]] == "B" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "C"
data_final[data_final[["Sample"]] == "MU-107" & data_final[["Location"]] == "D" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "C"
data_final[data_final[["Sample"]] == "MU-111" & data_final[["Location"]] == "B" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "E"
data_final[data_final[["Sample"]] == "MU-112" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "E"
data_final[data_final[["Sample"]] == "MU-199" & data_final[["Location"]] == "A" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "E"
data_final[data_final[["Sample"]] == "MU-199" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "3", "Usewear.type"] <- "C"
data_final[data_final[["Sample"]] == "MU-202" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "A"
data_final[data_final[["Sample"]] == "MU-202" & data_final[["Location"]] == "D" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "C"
data_final[data_final[["Sample"]] == "MU-224" & data_final[["Location"]] == "D" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "B2"
data_final[data_final[["Sample"]] == "MU-224" & data_final[["Location"]] == "B" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "C"
data_final[data_final[["Sample"]] == "MU-232" & data_final[["Location"]] == "B" &
             data_final[["Sublocation"]] == "2" & data_final[["Area"]] 
           == "01", "Usewear.type"] <- "E"
data_final[data_final[["Sample"]] == "MU-232" & data_final[["Location"]] == "B" &
             data_final[["Sublocation"]] == "2" & data_final[["Area"]] 
           == "02", "Usewear.type"] <- "A"
data_final[data_final[["Sample"]] == "MU-197" & data_final[["Location"]] == "B" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "F"
data_final[data_final[["Sample"]] == "MU-197" & data_final[["Location"]] == "D" &
             data_final[["Sublocation"]] == "3", "Usewear.type"] <- "F"
data_final[data_final[["Sample"]] == "MU-240" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "C+A"
data_final[data_final[["Sample"]] == "MU-240" & data_final[["Location"]] == "D" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "F"
data_final[data_final[["Sample"]] == "MU-246" & data_final[["Location"]] == "D" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "C/E"
data_final[data_final[["Sample"]] == "MU-273" & data_final[["Location"]] == "B" &
             data_final[["Sublocation"]] == "3", "Usewear.type"] <- "C/E"
data_final[data_final[["Sample"]] == "MU-019" & data_final[["Location"]] == "A" &
             data_final[["Sublocation"]] == "3", "Usewear.type"] <- "B2"
data_final[data_final[["Sample"]] == "MU-025" & data_final[["Location"]] == "B" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "B"
data_final[data_final[["Sample"]] == "MU-279" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "C"
data_final[data_final[["Sample"]] == "MU-104" & data_final[["Location"]] == "D" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "C/E"
data_final[data_final[["Sample"]] == "MU-119" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "C"
data_final[data_final[["Sample"]] == "MU-217" & data_final[["Location"]] == "A" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "A"

# add the use-wear type - Ramioul 
data_final[data_final[["Sample"]] == "R-002" & data_final[["Location"]] == "A" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "D"
data_final[data_final[["Sample"]] == "R-006" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "3", "Usewear.type"] <- "E"
data_final[data_final[["Sample"]] == "R-007" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "G"
data_final[data_final[["Sample"]] == "R-008" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "D"
data_final[data_final[["Sample"]] == "R-020" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "A"
data_final[data_final[["Sample"]] == "R-020" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "3", "Usewear.type"] <- "G"
data_final[data_final[["Sample"]] == "R-010" & data_final[["Location"]] == "B" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "C"
data_final[data_final[["Sample"]] == "R-010" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "C/E"
data_final[data_final[["Sample"]] == "R-013" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "G"
data_final[data_final[["Sample"]] == "R-013" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "3", "Usewear.type"] <- "D2"

# add the use-wear type - Buhlen
data_final[data_final[["Sample"]] == "BU-003" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "1", "Usewear.type"] <- "C"
data_final[data_final[["Sample"]] == "BU-032" & data_final[["Location"]] == "B" &
             data_final[["Sublocation"]] == "3", "Usewear.type"] <- "D2"
data_final[data_final[["Sample"]] == "BU-077" & data_final[["Location"]] == "B" &
             data_final[["Sublocation"]] == "3", "Usewear.type"] <- "C/E"
data_final[data_final[["Sample"]] == "BU-173" & data_final[["Location"]] == "C" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "C"
data_final[data_final[["Sample"]] == "BU-128" & data_final[["Location"]] == "D" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "C"
data_final[data_final[["Sample"]] == "BU-115" & data_final[["Location"]] == "A" &
             data_final[["Sublocation"]] == "2", "Usewear.type"] <- "C"
data_final[["Usewear.type"]] <- factor(data_final[["Usewear.type"]])

# correct information about the used objectives 
data_final[data_final[["Objective"]] == "50x09", "Objective"] <- "50x095"
data_final[data_final[["Objective"]] == "50x07", "Objective"] <- "50x075"
data_final[["Objective"]] <- factor(data_final[["Objective"]])

```


## Ignore some columns and reorder columns
```{r}
data_final <- data_final[c(1,54, 56, 55, 3:6, 57, 2, 7:53 )]

```


## Add units as comment()
```{r}
comment(data_final) <- units_var

```

Type `comment(data_final)` to check the units of the columns.


## Check the result
```{r}
str(data_final)
head(data_final)

```


---


# Save data
## Format name of output file
```{r}
file_out <- "Use-wear"

```
The files will be saved as "`r paste0("~/", file_out, ".[ext]")`".


## Write to XLSX
```{r}
write.xlsx(list(data = data_final, units = units_var_table), 
           file = paste0(dir_out, file_out, ".xlsx"))

```


## Save R object
```{r}
saveObject(data_final, file = paste0(dir_out, file_out, ".Rbin"))

```


---

# sessionInfo() and RStudio version

```{r}
sessionInfo()

```

RStudio version `r readLines("analysis/scripts/RStudioVersion.txt", n = 1)`.


---


END OF SCRIPT
